Viral Sequence Sorting Tools Evaluation

### 2022-10-05 EnvironmentalData/TARAMetagenomes/identifying_viral_contigs_environmental_tara.Rmd
Added lake michigan and TARA identifying viral contigs directories

```{r inputs}
# dataset name for file organization and outputs for metrics file
dataset_name <- "lake_michigan"
# checkV
checkV_path <- "merged_outputs/merged_checkv_assemblies.tsv"
# VIBRANT
vibrant_path <- "merged_outputs/merged_vibrant_assemblies.tsv"
# DeepVirFinder
dvf_path <- "merged_outputs/merged_dvf_assemblies.tsv"
# Virsorter
vs_path <- "merged_outputs/merged_virsorter_assemblies.tsv"
# Virsorter2
vs2_path <- "merged_outputs/merged_vs2_assemblies.tsv"
# Kaiju
kj_path <- "merged_outputs/merged_kaiju_assemblies.tsv"
# KB cutoff
KB_CUTOFF <- 3000

@@

## checkV
```{r}
checkV <- fread(checkV_path, 
                sep="\t",
                header = T, 
                select = c(
                    'contig_id',
                    'provirus',
                    'completeness',
                    'contamination',
                    'viral_genes',
                    'host_genes',
                    'gene_count',
                    'contig_length',
                    'checkv_quality'
                    )
                ) %>% 
    rename(
        contig = contig_id,
        checkv_provirus = provirus,
        checkv_completeness = completeness,
        checkv_contamination = contamination,
        checkv_viral_genes = viral_genes,
        checkv_host_genes = host_genes,
        checkv_total_genes = gene_count,
        checkv_length = contig_length
        )
checkV$method = 'checkv'
checkV$contig <- gsub('_', '', checkV$contig)
```

## VIBRANT
```{r}
vb_c <- fread(vibrant_path,
              header = T,
              sep = "\t",
              select = c(
                  'scaffold',
                  'type',
                  'Quality'
                  )
              ) %>%
    rename(
        contig = scaffold,
        vibrant_quality = Quality
        )
vb_c$method <- "vibrant"
vb_c$vibrant_prophage <- "No"
vb_c$vibrant_prophage[grep("_fragment_", vb_c$contig)] <- "Yes"
vb_c$contig <- gsub("_fragment_.*", "", vb_c$contig)
vb_c$contig <- gsub('_', '', vb_c$contig)
```

## DeepVirFinder
```{r}
dvf_c <- fread(dvf_path,
               header = T,
               sep = "\t",
               select = c(
                   'name',
                   'assembly',
                   'score',
                   'pvalue'
                   )
               ) %>% 
    rename(
        contig = name
        )
dvf_c$contig <- gsub('_', '', dvf_c$contig)
dvf_c$bh_pvalue <- p.adjust(dvf_c$pvalue, method = "BH")
```

## VirSorter
```{r}
vs_c <- fread(vs_path,
              select = c(
                  '## Contig_id',
                  'Nb genes contigs',
                  'Nb genes',
                  'Category',
                  'Nb phage hallmark genes'
                  )
              ) %>% drop_na(`## Contig_id`)
vs_c$`## Contig_id` <- sub("VIRSorter_", "", vs_c$`## Contig_id`)
vs_c$`## Contig_id` <- sub("-circular", "", vs_c$`## Contig_id`)
vs_c$`## Contig_id` <- gsub("_", "", vs_c$`## Contig_id`)
vs_c <- separate(vs_c, col = `## Contig_id`, into = c("assembly", "contig"), sep="--")

# convert to numeric
vs_c$category <- as.numeric(vs_c$Category)
vs_c <- vs_c %>% drop_na(category)
```

## VirSorter2 
```{r}
vs2_c <- fread(vs2_path,
                    header = T,
                    sep = '\t',
                    select = c(
                        'seqname',
                        'dsDNAphage',
                        'RNA',
                        'ssDNA',
                        'max_score',
                        'max_score_group',
                        'hallmark',
                        'viral',
                        'cellular'
                    )
               ) %>% 
              separate(
                  col = seqname,
                  into = c("contig", "type"), 
                  sep = "\\|\\|",
                  remove = T
                  )
vs2_c$contig <- gsub("_", "", vs2_c$contig)
```

@@

colnames(kj_c) <-c(
    "Classified",
    "Contig",
    "NCBI_taxon",
    "len",
    "ID_best",
    "IDs_all",
    "Seq",
    "Name",
    "assembly"
    )
kj_c$Contig <- gsub('_', '', kj_c$Contig)
kj_c <- setDT(kj_c)[, c("assembly", "contig") := tstrsplit(Contig, "--")]
kj_c <- setDT(kj_c)[, c("Kaiju_Viral", "Kingdom") := tstrsplit(Name, ";", keep = c(1, 2))]
```

@@

```{r}
viruses <- full_join(x=vs2_c, y=dvf_c, by = "contig")
viruses <- full_join(x=viruses, y=checkV, by = "contig")
viruses <- full_join(x=viruses, y=vb_c, by = "contig")
viruses <- full_join(x=viruses, y=vs_c, by = "contig")
viruses <- full_join(x=viruses, y=kj_c, by = "contig")
```

Remove contigs not greater than the basepair length cutoff (3000)

@@

#Virsorter
viruses$category[is.na(viruses$category)] <- 0

@@

  if (include_virsorter) {
    keep_score[input_seqs$category==1] <- keep_score[input_seqs$category==1] + 1
    keep_score[input_seqs$category==2] <- keep_score[input_seqs$category==2] + 1
    keep_score[input_seqs$category==3] <- keep_score[input_seqs$category==3] + 0.5
    keep_score[input_seqs$category==4] <- keep_score[input_seqs$category==4] + 1
    keep_score[input_seqs$category==5] <- keep_score[input_seqs$category==5] + 1
    keep_score[input_seqs$category==6] <- keep_score[input_seqs$category==6] + 0.5
  }

@@

```{r}
# histogram of contig lengths
ggplot(viruses, aes(x=checkv_length)) + geom_histogram(binwidth=10000)
```

>>>>

```{r inputs}
# dataset name for file organization and outputs for metrics file
dataset_name <- "tara"
# checkV
checkV_path <- "merged_outputs/merged_checkv_assemblies.tsv"
# VIBRANT
vibrant_path <- "merged_outputs/merged_vibrant_assemblies.tsv"
# DeepVirFinder
dvf_path <- "merged_outputs/merged_dvf_assemblies.tsv"
# Virsorter
vs_path <- "merged_outputs/merged_virsorter_assemblies.tsv"
# Virsorter2
vs2_path <- "merged_outputs/merged_vs2_assemblies.tsv"
# Kaiju
kj_path <- "merged_outputs/merged_kaiju_assemblies.tsv"
# KB cutoff
KB_CUTOFF <- 3000

@@

## checkV
```{r}
checkV <- fread(checkV_path, 
                sep="\t",
                header = T, 
                select = c(
                    'contig_id',
                    'provirus',
                    'completeness',
                    'contamination',
                    'viral_genes',
                    'host_genes',
                    'gene_count',
                    'contig_length',
                    'checkv_quality',
                    'assembly'
                    )
                ) %>% 
    rename(
        contig = contig_id,
        checkv_provirus = provirus,
        checkv_completeness = completeness,
        checkv_contamination = contamination,
        checkv_viral_genes = viral_genes,
        checkv_host_genes = host_genes,
        checkv_total_genes = gene_count,
        checkv_length = contig_length
        )
checkV$method = 'checkv'
checkV$contig <- gsub('_', '', checkV$contig)
checkV$uniq_contig <- paste(checkV$assembly, checkV$contig, sep = "--")
```

## VIBRANT
```{r}
vb_c <- fread(vibrant_path,
              header = T,
              sep = "\t",
              select = c(
                  'scaffold',
                  'assembly',
                  'type',
                  'Quality'
                  )
              ) %>%
    rename(
        contig = scaffold,
        vibrant_quality = Quality
        )
vb_c$method <- "vibrant"
vb_c$vibrant_prophage <- "No"
vb_c$vibrant_prophage[grep("_fragment_", vb_c$contig)] <- "Yes"
vb_c$contig <- gsub("_fragment_.*", "", vb_c$contig)
vb_c$contig <- gsub('_', '', vb_c$contig)
vb_c$uniq_contig <- paste(vb_c$assembly, vb_c$contig, sep = "--")
```

## DeepVirFinder
```{r}
dvf_c <- fread(dvf_path,
               header = T,
               sep = "\t",
               select = c(
                   'name',
                   'assembly',
                   'score',
                   'pvalue'
                   )
               ) %>% 
    rename(
        contig = name
        )
dvf_c$contig <- gsub('_', '', dvf_c$contig)
dvf_c$bh_pvalue <- p.adjust(dvf_c$pvalue, method = "BH")
dvf_c$uniq_contig <- paste(dvf_c$assembly, dvf_c$contig, sep = "--")
```

## VirSorter
```{r}
vs_c <- fread(vs_path,
              select = c(
                  '## Contig_id',
                  'Nb genes contigs',
                  'Nb genes',
                  'Category',
                  'Nb phage hallmark genes'
                  )
              ) %>% drop_na(`## Contig_id`)
vs_c$`## Contig_id` <- sub("VIRSorter_", "", vs_c$`## Contig_id`)
vs_c$`## Contig_id` <- sub("-circular", "", vs_c$`## Contig_id`)
vs_c$`## Contig_id` <- gsub("_", "", vs_c$`## Contig_id`)
vs_c <- vs_c %>% rename(uniq_contig = `## Contig_id`)
# convert to numeric
vs_c$Category <- as.numeric(vs_c$Category)
vs_c <- vs_c %>% drop_na(Category)
```

## VirSorter2 
```{r}
vs2_c <- fread(vs2_path,
                    header = T,
                    sep = '\t',
                    select = c(
                        'seqname',
                        'dsDNAphage',
                        'RNA',
                        'ssDNA',
                        'max_score',
                        'max_score_group',
                        'hallmark',
                        'viral',
                        'cellular',
                        'assembly'
                    )
               ) %>% 
              separate(
                  col = seqname,
                  into = c("contig", "type"), 
                  sep = "\\|\\|",
                  remove = T
                  )
vs2_c$contig <- gsub("_", "", vs2_c$contig)
vs2_c$uniq_contig <- paste(vs2_c$assembly, vs2_c$contig, sep="--")
```

@@

colnames(kj_c) <-c(
    "Classified",
    "uniq_contig",
    "NCBI_taxon",
    "len",
    "ID_best",
    "IDs_all",
    "Seq",
    "Name",
    "assembly"
    )
kj_c$Contig <- gsub('_', '', kj_c$Contig)
kj_c <- setDT(kj_c)[, c("Kaiju_Viral", "Kingdom") := tstrsplit(Name, ";", keep = c(1, 2))]
```

@@

```{r}
viruses <- full_join(x=vs2_c, y=dvf_c, by = "uniq_contig")
viruses <- full_join(x=viruses, y=checkV, by = "uniq_contig")
viruses <- full_join(x=viruses, y=vb_c, by = "uniq_contig")
viruses <- full_join(x=viruses, y=vs_c, by = "uniq_contig")
viruses <- full_join(x=viruses, y=kj_c, by = "uniq_contig")
```

Remove contigs greater than the basepair length cutoff (3000)

@@

#Virsorter
viruses$Category[is.na(viruses$Category)] <- 0

@@

  if (include_virsorter) {
    keep_score[input_seqs$Category==1] <- keep_score[input_seqs$Category==1] + 1
    keep_score[input_seqs$Category==2] <- keep_score[input_seqs$Category==2] + 1
    keep_score[input_seqs$Category==3] <- keep_score[input_seqs$Category==3] + 0.5
    keep_score[input_seqs$Category==4] <- keep_score[input_seqs$Category==4] + 1
    keep_score[input_seqs$Category==5] <- keep_score[input_seqs$Category==5] + 1
    keep_score[input_seqs$Category==6] <- keep_score[input_seqs$Category==6] + 0.5
  }

@@

```{r}
# histogram of contig lengths
ggplot(viruses, aes(x=checkv_length)) + 
    geom_histogram(bins=1000) +
    scale_y_log10()
ggplot(checkV, aes(x=checkv_length)) + 
    geom_histogram(bins=1000) +
    scale_y_log10()

```{r}
# histogram of contig lengths
ggplot(checkV, aes(x=checkv_length)) + 
    geom_histogram(bins=1000) +
    scale_y_log10() +
```
